Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
finished a round of specific comments
WML: first round of edits complete.
created (most of) the supplementary materials
post will's comments

% WML: "borrowing information" is intuitive and informal. I agree that the term should be used, but there might be a slightly more appropriate use. Not a big deal.

@@

% WML: We were debating whether to use g for gene or f for feature, and either choice would work. But we use g, we should probably also use G to denote the number of genes, rather than F.

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,F$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
\label{e:data}
\end{equation}

% WML: Are the sequencing depths c_{vi} or \delta_{vi}? I think c_{vi} is better because \delta is used in the phi-alpha-delta parameterization of gene expression effects.

@@

% WML: Is \mu_g the vector of \mu_{gv} values? If so, please clarify.

We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations \citep{gelman2006prior}, e.g. $\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution. In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.

% WML: I assume the "hierarchical means" are the hyperparameters in the \eta vector and the "marginally uniform correlations" are the ones derived from the off-diagonal covariance elements of \Sigma. Small clarifications might be helpful in passing.

@@

% WML: The use of \phi in this section conflicts with the phi-alpha-delta parameterization in other heterosis models. Maybe you could replace phi, alpha, and delta by other letters in the ShrinkBayes section to avoid this conflict.

@@

% WML: \ref{s:intro} or \ref{s:heterosis}?

% WML: the first two sentences are unclear. I would delete the first sentence altogether and compare "varieties" to treatment groups or blocks for the sake of clarity.

As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$,

% WML: For mid-parent heterosis, you need \mu_{g3} \ne \frac{\mu_{g1} + \mu_{g2}}{2}. I don't know whether you need \mu_{g1}\le \mu_{g3}\le \mu_{g2} or \mu_{g2}\le \mu_{g3}\le \mu_{g1}, and I have doubts. It might be good to check with Dan on this.

mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.

% WML: You may want to clarify that the genes in the list are ranked by the strength of EH and then justify the lists's utility for geneticists and in the comparison study to follow. This could be a good segue.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. We compared our approach to approaches using {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

% WML: Please clarify that {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} are R packages. You may even want to cite R somewhere.

@@

% WML: Sentences 2-4 in the paragraph below are redundant with sentence 1. I commented out the paragraph below and inserted a modified one afterwards
%To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means. We estimated sample-specific normalization factors to account for sequencing depth. Then, we estimated common a common dispersion for each gene, but shared across the samples. Finally, we obtained independent estimates for the gene-variety specific location parameters. The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study, i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.

To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated all parameters as truth for the simulation study:  i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide statistics to measure the strength of heterosis of each gene.

% WML: "... provide a statistic representing the plausibility of heterosis, i.e. higher values indicating higher plausibility." => "...provide statistics to measure the strength of heterosis of each gene."

% WML: I rewrote the paragraph on edgeR.

@@

% NEED MLE

% WML: I rewrote the baySeq paragraph.

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}.

% WML: your use of "following" in this last sentence is not clear to me.

% WML: Here, you use the phi-alpha-delta parameterization. You already defined phi and delta in previous sections. To avoid a conflict, you could easily change the notation here. It doesn't even have to be super intuitive because you only use it in this one paragraph.

 Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$. Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.

@@

For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see the approaches based on the model in Section \ref{s:model}, i.e. both RStan approaches and the ShrinkBayes approach, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases our ability to distinguish genes with heterosis from genes without improves.

@@

%\input{stan_models}

>>>>

@@

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
\label{e:data}
\end{equation}

@@

We assume improper uniform priors for $\eta$ and $\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution. In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.

@@

@@

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
\label{e:hypotheses}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis. This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

@@

To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated these parameters as the truth for the simulation study: i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide statistics to measure the strength of extreme heterosis of each gene.

@@

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\gamma_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\zeta_g = \mu_{g3} - \gamma_g$ be the hybrid effect. After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\zeta}_g$ and $\hat{\alpha}_g$ of $\zeta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$. Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\zeta}_g| \le |\hat{\alpha}_g|$, $P(\zeta_g + \alpha_g > 0 |y)$ if $\hat{\zeta}_g > -\hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g > 0 |y)$ if $\hat{\zeta}_g > \hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g < 0 |y)$ if $\hat{\zeta}_g < \hat{\alpha}_g \le 0$, and $P(\zeta_g + \alpha_g < 0 |y)$ if $\hat{\zeta}_g < -\hat{\alpha}_g \le 0$.

@@

For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}, \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see the approaches based on the model in Section \ref{s:model}, i.e. both RStan approaches and the ShrinkBayes approach, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases our ability to distinguish genes with heterosis from genes without improves.

@@
